A sheaf--function dictionary for algebraic tori over non-archimedean local fields

### 2013-08-16 shanghai.tex
Improvements to the first two sections of Part 2.

\begin{proposition}\label{prop:functorialG}
Trace of Frobenius $\QCiso{G} \to G(\Fq)$ is a natural transformation 
between the following two additive functors from the category of smooth commutative
group schemes locally of finite type over $\Fq$ to the category of abelian groups:
\begin{enumerate}
\item[$F_1$:] $G \mapsto \QCiso{G}$;
\item[$F_2$:] $G \mapsto G(\Fq)$.
\end{enumerate}
\end{proposition}

@@

  To show that $L_G(f,\psi) \iso (\cs{L}_0,\mu_0)$ we simply apply the functions--sheaves dictionary
  as follows ({\it cf} \cite{laumon:87a}*{1.1.3} or, ultimately, \cite{deligne:SGA4.5}*{Sommes trigonométriques}).
  First, recall that $L_G(f,\psi) = (L(f,\psi), \mu(f,\psi))$. 
  By  \cite{laumon:87a}*{1.1.3.3}, $\trFrob{(G_{k'}\hookrightarrow G)^*L(f,\psi)} = \trFrob{\cs{L}} \circ N_{k'/k}$, for each finite extension $k'/k$.
  On the other hand, $\trFrob{\cs{L}} \circ N_{k'/k} = \trFrob{(G_{k'}\hookrightarrow G)^*\cs{L}}$,  by Proposition~\ref{prop:basechange}.
  %Using notation from \cite{laumon:87a}*{1.1}, this means $t_{\bullet L(f,\psi)} = t_{\bullet \cs{L}_0}$.
  Using \cite{laumon:87a}*{Th\'eor\`eme~1.1.2}, it follows that  $L(f,\psi) \iso \cs{L}_0$ in $\Loc(G)$.
  It is now clear that $(L(f,\psi),\mu(f,\psi)) \iso (\cs{L}_0,\mu_0)$ in $\QCb(G)$.
  This completes the proof that $L_G : \QCf(G) \to \QC(G)$ is essentially surjective.
  Since we have already seen (in Section~\ref{sec:finite}) that $L_G$ is full and faithful, 
  it follows that $L_G$ is an equivalence.

@@

 \labitem{(cs.1)}{cs.1} an indexed set of isomorphisms
  $\mu_{x,y} : \gcs{L}_{x+y} \mathop{\longrightarrow}\limits^{\iso} \gcs{L}_{x} \otimes_{\EE}\gcs{L}_{y}$,
  for all $x,y \in G(\bFq)$, such that
  \[
   \begin{tikzcd}[row sep=40]
    \gcs{L}_{x+y+z} \arrow{rr}{\mu_{x+y,z}} \arrow[swap]{d}{\mu_{x,y+z}}
    && \gcs{L}_{x+y}\otimes_{\EE}\gcs{L}_{z} \dar{\mu_{x,y} \tight{0.5}{\otimes}{1} \id} \\
    \gcs{L}_{x} \otimes_{\EE}\gcs{L}_{y+z} \arrow{rr}{\id \otimes_{\EE}\mu_{y,z}}
    && \gcs{L}_{x} \otimes\gcs{L}_{y} \otimes_{\EE}\gcs{L}_{z}
   \end{tikzcd}
  \]
  commutes, for all $x,y,z\in G(\bFq)$; and

 \labitem{(cs.2)}{cs.2} an indexed set of isomorphisms $\phi_{x} : \gcs{L}_{\Frob{\Fq}(x)} \to \gcs{L}_x$
  such that
  \[
   \begin{tikzcd}[row sep=40]
    \gcs{L}_{\Frob{\Fq}(x)+\Frob{\Fq}(y)} \arrow[swap]{d}{\phi_{x+y}} \arrow{rr}{\mu_{\Frob{\Fq}(x),\Frob{\Fq}(y)}}
    && \gcs{L}_{\Frob{\Fq}(x)}\otimes_{\EE}\gcs{L}_{\Frob{\Fq}(y)} \dar{\phi_x \tight{0}{\otimes}{0} \phi_y} \\
    \gcs{L}_{x+y} \arrow{rr}{\mu_{x,y}}
    && \gcs{L}_x \otimes\gcs{L}_y
   \end{tikzcd}
  \]
  commutes, for all $x,y\in G(\bFq)$.
\end{enumerate}
Under this equivalence, a morphism $\alpha : \cs{L} \to \cs{L'}$ of quasicharacter sheaves on $G$ is given by 
\begin{enumerate}
 \labitem{(cs.3)}{cs.3} an indexed set $\alpha_x : \gcs{L}_x \to \gcs{L'}_x$
  of linear transformations (of one-dimensional $\EE$-vector spaces),
  as $x$ runs over $G(\bFq)$, such that
  \[
   \begin{tikzcd}[column sep=40]
    \arrow[swap]{d}{\phi_x} \gcs{L}_{\Frob{\Fq}(x)} \rar{\alpha_{\Frob{\Fq}(x)}} & \gcs{L'}_{\Frob{\Fq}(x)} \dar{\phi_x'}
    &\arrow[draw=none]{d}[pos=.4,description]{\text{\normalsize{and}}}& \arrow[swap]{d}{\mu} \gcs{L}_{x+y} \rar{\alpha_{x+y}} & \gcs{L'}_{x+y} \dar{\mu'_{x+y}} \\
    \gcs{L}_x \rar{\alpha_x} & \gcs{L'}_x
    & {} & \gcs{L}_x\otimes\gcs{L}_y \rar{\alpha_x\otimes_{\EE}\alpha_y} & \gcs{L'}_x \otimes_{\EE}\gcs{L'}_y
   \end{tikzcd}
  \]
  both commute.
\end{enumerate}

@@

  There is a unique isomorphism $\gcs{L}_0 \cong \EE$ under which $\mu_{0,0}$ corresponds
  to the multiplication map $\EE \otimes_{\EE}\EE \to \EE$.
  The condition $\mu_{0,0}(s(0)) = s(0) \otimes_{\EE}s(0)$ together with the non-triviality of $s$
  then forces $s(0) = 1$.

@@

A N\'eron model for $T$, denoted here by $\TT$, 
is a separated, smooth, commutative group scheme locally of finite type over $\OK$ with generic fibre $T$, that enjoys the N\'eron mapping property: \Clifton{Include reference to BLR.}
if $S$ is a smooth scheme over $\OK$ then every morphism $S_\eta \to T$
over $K$ extends uniquely to a morphism $S \to \TT$. Equivalently, the
N\'eron mapping property states that the function
$\Hom_{\OK} (S,\TT) \to \Hom_K(S_\eta,T)$, given by restriction to
generic fibres, is a bijection; 
in particular, taking $S = \Spec{\OK}$, this implies $\TT(\OK) = T(K)$.

@@

\begin{theorem}\label{thm:application}
Let $d$ be a positive integer.
\begin{enumerate}
\item 
For every algebraic torus $T$ over $K$, the trace of Frobenius 
  \[
\trFrob{T,d}:  \QCiso{\GN{T}_d} \to \Hom_{< d}(T(K),\EEx)
  \]
is an isomomorphism. 
\item The isomorphism $\trFrob{T,d}$ is functorial in the following sense: $T \mapsto \trFrob{T,d}$ is a natural transformation
from the additive functor
\[T \mapsto \QCiso{\GN{T}_d}\]
to the additive functor
\[
T \mapsto \Hom_{<d}(T(K),\EEx).
\]
\end{enumerate}
\end{theorem}

>>>>

\begin{proposition}\label{prop:functorialG}
Trace of Frobenius $\QCiso{G} \to G(\Fq)$ is a natural transformation 
between the following two additive functors 
\begin{enumerate}
\item[$F_1$:] $G \mapsto \QCiso{G}$
\item[$F_2$:] $G \mapsto G(\Fq)$
\end{enumerate}
from the category of smooth commutative
group schemes locally of finite type over $\Fq$ to the category of abelian groups.
\end{proposition}

@@

  To show that $L_G(f,\psi) \iso (\cs{L}_0,\mu_0)$ we simply apply the functions--sheaves dictionary
  as follows ({\it cf} \cite{laumon:87a}*{1.1.3} or, ultimately, \cite{deligne:SGA4.5}*{Sommes trigonométriques}).
  First, recall that $L_G(f,\psi) = (L(f,\psi), \mu(f,\psi))$. 
  By  \cite{laumon:87a}*{1.1.3.3}, 
  \[
  \trFrob{(G_{k'}\hookrightarrow G)^*L(f,\psi)} = \trFrob{\cs{L}} \circ N_{k'/k},
  \]
   for each finite extension $k'/k$.
  On the other hand, 
  \[
  \trFrob{\cs{L}} \circ N_{k'/k} = \trFrob{(G_{k'}\hookrightarrow G)^*\cs{L}},
  \]
   by Proposition~\ref{prop:basechange}.
  %Using notation from \cite{laumon:87a}*{1.1}, this means $t_{\bullet L(f,\psi)} = t_{\bullet \cs{L}_0}$.
  Using \cite{laumon:87a}*{Th\'eor\`eme~1.1.2}, it follows that  $L(f,\psi) \iso \cs{L}_0$ in $\Loc(G)$.
  It is now clear that $(L(f,\psi),\mu(f,\psi)) \iso (\cs{L}_0,\mu_0)$ in $\QCb(G)$.
  This completes the proof that $L_G : \QCf(G) \to \QC(G)$ is essentially surjective.
  Since we have already seen (in Section~\ref{sec:finite}) that $L_G$ is full and faithful, 
  it follows that $L_G$ is an equivalence.

@@

 \labitem{(cs.1)}{cs.1} an indexed set of isomorphisms
  $\mu_{x,y} : \gcs{L}_{x+y} \mathop{\longrightarrow}\limits^{\iso} \gcs{L}_{x} \otimes_{\EE}\gcs{L}_{y}$,
  for all $x,y \in G(\bFq)$, such that
  \[
   \begin{tikzcd}[row sep=40]
    \gcs{L}_{x+y+z} \arrow{rr}{\mu_{x+y,z}} \arrow[swap]{d}{\mu_{x,y+z}}
    && \gcs{L}_{x+y}\otimes_{\EE}\gcs{L}_{z} \dar{\mu_{x,y} \tight{0.5}{\otimes}{1} \id} \\
    \gcs{L}_{x} \otimes_{\EE}\gcs{L}_{y+z} \arrow{rr}{\id \otimes\mu_{y,z}}
    && \gcs{L}_{x} \otimes\gcs{L}_{y} \otimes_{\EE}\gcs{L}_{z}
   \end{tikzcd}
  \]
  commutes, for all $x,y,z\in G(\bFq)$; and

 \labitem{(cs.2)}{cs.2} an indexed set of isomorphisms $\phi_{x} : \gcs{L}_{\Frob{\Fq}(x)} \to \gcs{L}_x$
  such that
  \[
   \begin{tikzcd}[row sep=40]
    \gcs{L}_{\Frob{\Fq}(x)+\Frob{\Fq}(y)} \arrow[swap]{d}{\phi_{x+y}} \arrow{rr}{\mu_{\Frob{\Fq}(x),\Frob{\Fq}(y)}}
    && \gcs{L}_{\Frob{\Fq}(x)}\otimes_{\EE}\gcs{L}_{\Frob{\Fq}(y)} \dar{\phi_x \tight{0}{\otimes}{0} \phi_y} \\
    \gcs{L}_{x+y} \arrow{rr}{\mu_{x,y}}
    && \gcs{L}_x \otimes\gcs{L}_y
   \end{tikzcd}
  \]
  commutes, for all $x,y\in G(\bFq)$.
\end{enumerate}
Under this equivalence, a morphism $\alpha : \cs{L} \to \cs{L'}$ of quasicharacter sheaves on $G$ is given by 
\begin{enumerate}
 \labitem{(cs.3)}{cs.3} an indexed set $\alpha_x : \gcs{L}_x \to \gcs{L'}_x$
  of linear transformations (of one-dimensional $\EE$-vector spaces),
  as $x$ runs over $G(\bFq)$, such that
  \[
   \begin{tikzcd}[column sep=40]
    \arrow[swap]{d}{\phi_x} \gcs{L}_{\Frob{\Fq}(x)} \rar{\alpha_{\Frob{\Fq}(x)}} & \gcs{L'}_{\Frob{\Fq}(x)} \dar{\phi_x'}
    &\arrow[draw=none]{d}[pos=.4,description]{\text{\normalsize{and}}}& \arrow[swap]{d}{\mu} \gcs{L}_{x+y} \rar{\alpha_{x+y}} & \gcs{L'}_{x+y} \dar{\mu'_{x+y}} \\
    \gcs{L}_x \rar{\alpha_x} & \gcs{L'}_x
    & {} & \gcs{L}_x\otimes\gcs{L}_y \rar{\alpha_x\otimes\alpha_y} & \gcs{L'}_x \otimes_{\EE}\gcs{L'}_y
   \end{tikzcd}
  \]
  both commute.
\end{enumerate}

@@

  There is a unique isomorphism $\gcs{L}_0 \cong \EE$ under which $\mu_{0,0}$ corresponds
  to the multiplication map $\EE \otimes_{\EE}\EE \to \EE$.
  The condition $\mu_{0,0}(s(0)) = s(0) \otimes s(0)$ together with the non-triviality of $s$
  then forces $s(0) = 1$.

@@

A N\'eron model for $T$, denoted here by $\TT$ 
and whose existence is given by \cite{bosch-lutkebohmert-reynaud:NeronModels}*{Ch 10.1, Thm 6},
is a separated, smooth, commutative group scheme locally of finite type over $\OK$ with generic fibre $T$, that enjoys the N\'eron mapping property:
if $S$ is a smooth scheme over $\OK$ then every morphism $S_\eta \to T$
over $K$ extends uniquely to a morphism $S \to \TT$. 
Equivalently, the
N\'eron mapping property states that the function
$\Hom_{\OK} (S,\TT) \to \Hom_K(S_\eta,T)$, given by restriction to
generic fibres, is a bijection; 
in particular, taking $S = \Spec{\OK}$, this implies $\TT(\OK) = T(K)$.

@@

\begin{theorem}\label{thm:application}
Let $d$ be a positive integer.
For every algebraic torus $T$ over $K$, the trace of Frobenius 
  \[
\trFrob{T,d}:  \QCiso{\GN{T}_d} \to \Hom_{< d}(T(K),\EEx)
  \]
is an isomomorphism. 
Moreover, the isomorphism $\trFrob{T,d}$ is functorial in the following sense: $T \mapsto \trFrob{T,d}$ is a natural transformation between the following two additive functors
\begin{enumerate}
\item[$F_1$:] $T \mapsto \QCiso{\GN{T}_d}$
\item[$F_2$:] $T \mapsto \Hom_{<d}(T(K),\EEx)$
\end{enumerate}
from the additive category of algebraic tori over $K$ to the category of abelian groups.
\end{theorem}
